---
title: 'stars: data model'
author: "Edzer Pebesma"
date: "September 28, 2018"
output:
  html_document:
    toc: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Stars: data model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(collapse = TRUE)
set.seed(13579)
```

This vignette explains the data model of `stars` objects, illustrated using artificial and real datasets.

## Grid type

### Regular grids
With a very simple file created from a $4 \times 5$ matrix
```{r fig.width=4.5, fig.height=4}
suppressPackageStartupMessages(library(stars))
m = matrix(1:20, nrow = 5, ncol = 4)
dim(m) = c(x = 5, y = 4) # named dim
(s = st_as_stars(m))
```
we see that

* the rows (5) are mapped to the first dimension, the x-coordinate
* the columns (4) are mapped to the second dimension, the y-coordinate
* the `from` and `to` fields of each dimension define a range that corresponds to the array dimension:

```{r}
dim(s[[1]])
```
* offset and delta specify how increasing row and column index maps to x and y coordinate values respectively.

When we plot this object, using the `image` method for `stars` objects, 
```{r fig.width=4.5, fig.height=4}
image(s, text_values = TRUE, axes = TRUE)
```

we see that $(0,0)$ is the origin of the grid (grid corner), and $1$ the coordinate value increase from one index (row, col) to the next. It means that consecutive matrix columns represent grid lines, going from south to north. Grids defined this way are **regular**: grid cell size is constant everywhere.

Many actual grid datasets have y coordinates (grid rows) going from North to South (top to bottom); this is realised with a negative value for `delta`. We see that the grid origing $(0,0)$ did not change:
```{r fig.width=4.5, fig.height=4}
attr(s, "dimensions")[[2]]$delta = -1
image(s, text_values = TRUE, axes = TRUE)
```

An example is the GeoTIFF carried in the package, which, as probably all data sources read through GDAL, has a negative `delta` for the `y`-coordinate:
```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
st_dimensions(read_stars(tif))["y"]
```

### Raster attributes, rotated and sheared grids

Dimension tables of `stars` objects carry a `raster` attribute:
```{r}
str(attr(st_dimensions(s), "raster"))
```

which is a list that holds

* `dimensions`: character, the names of raster dimensions (if any), as opposed to e.g. spectral, temporal or other dimensions
* `affine`: numeric, the affine parameters
* `curvilinear`: a logical indicating whether the raster is curvilinear

These fields are needed at this level, because they describe properties of the array at a higher level than individual dimensions do: a pair of dimensions forms a raster, both `affine` and `curvilinear` describe how x and y _as a pair_ are derived from grid indexes (see below) when this cannot be done on a per-dimension basis.

With two affine parameters $a_1$ and $a_2$, $x$ and $y$ coordinates are derived from (1-based) grid indexes $i$ and $j$, grid offset values $o_x$ and $o_y$, and grid cell sizes $d_x$ and $d_y$ by

$$x = o_x + (i-1) d_x + (j-1) a_1$$
    
$$y = o_y + (i-1) a_2 + (j-1) d_y$$
Clearly, when $a_1=a_2=0$, $x$ and $y$ are entirely derived from their respective index, offset and cellsize.

Note that for integer indexes, the coordinates are that of the starting edge of a grid cell; to get the grid cell center of the top left grid cell (in case of a negative $d_y$), use $i=1.5$ and $j=1.5$.

We can rotate grids by setting $a_1$ and $a_2$ to a non-zero value:

```{r}
attr(attr(s, "dimensions"), "raster")$affine = c(0.1, 0.1)
plot(st_as_sf(s, as_points = FALSE), axes = TRUE, nbreaks = 20)
```

The rotation angle, in degrees, is

```{r}
atan2(0.1, 1) * 180 / pi
```

### Regular grids: sheared, sheared and rotated

Sheared grids are obtained when the two rotation coefficients, $a_1$ and $a_2$, are unequal:
```{r}
attr(attr(s, "dimensions"), "raster")$affine = c(0.1, 0.2)
plot(st_as_sf(s, as_points = FALSE), axes = TRUE, nbreaks = 20)
```

Now, the y-axis and x-axis have different rotation in degrees of respectively
```{r}
atan2(c(0.1, 0.2), 1) * 180 / pi
```

## Rectilinear grids

[Rectilinear grids](https://en.wikipedia.org/wiki/Regular_grid) have orthogonal axes, but do not have congruent (equally sized and shaped) cells: each axes has its own irregular subdivision.

```{r}
x = c(0,0.5,1,2,4,5)
y = c(0.3,0.5,1,2,2.2)
r = st_as_stars(list(m = m), dimensions = st_dimensions(x = x, y = y))
r
#plot(st_as_sf(r, as_points = FALSE), axes = TRUE, nbreaks = 20)
image(x, y, m, col = sf.colors(21))
x1 = c(0.25,0.75,1.5,3,4.5)
y1 = c(0.4,0.75,1.5,2.1)
image(x1, y1, m, col = sf.colors(21))
```

## Curvilinear grids

Curvilinear grids are grids whose grid lines are not straight. Rather than describing the curvature parametrically, the typical (HDF5 or netcdf) files in which they are found have two raster layers, one with the latitudes and one with the longitudes for every grid cell. Using the GDAL interface, we need to compose a `stars` object from the array and a long and lat matrix. The matrices with longitude and latitude values are stored in the `values` field of their respective dimensions.

As an example, we will use a Sentinel 5P dataset available from package `starsdata`; this package can be installed with
```{r eval=FALSE}
install.packages("starsdata", repos = "http://pebesma.staff.ifgi.de", type = "source") 
```

The dataset is found here:
```{r}
(s5p = system.file("sentinel5p/S5P_NRTI_L2__NO2____20180717T120113_20180717T120613_03932_01_010002_20180717T125231.nc", package = "starsdata"))
```

```{r echo=FALSE}
EVAL = s5p != ""
```
We can construct the curvilinear `stars` raster by calling `st_as_stars` on three arrays, with data, longitude and latitudes:
```{r eval=EVAL}
lat_ds = paste0("HDF5:\"", s5p, "\"://PRODUCT/latitude")
lon_ds = paste0("HDF5:\"", s5p, "\"://PRODUCT/longitude")
nit_ds = paste0("HDF5:\"", s5p, "\"://PRODUCT/SUPPORT_DATA/DETAILED_RESULTS/nitrogendioxide_summed_total_column")
lat = read_stars(lat_ds)
lon = read_stars(lon_ds)
nit = read_stars(nit_ds)
nit[[1]][nit[[1]] > 9e+36] = NA

ll = setNames(c(lon, lat), c("x", "y"))
nit.c = st_as_stars(nit, curvilinear = ll)
st_crs(nit.c) = 4326
nit.c
```

```{r eval=EVAL}
plot(nit.c, breaks = "equal", reset = FALSE, axes = TRUE, as_points = TRUE, pch = 16, key.pos = NULL)
maps::map('world', add = TRUE, col = 'red')
```

```{r eval=EVAL}
plot(nit.c, breaks = "equal", reset = FALSE, axes = TRUE, as_points = FALSE, border = NA)
maps::map('world', add = TRUE, col = 'red')
```

We can downsample the data by
```{r eval=EVAL}
(nit.c = stars:::st_downsample(nit.c, 8))
plot(nit.c, breaks = "equal", reset = FALSE, axes = TRUE, as_points = TRUE, pch = 16)
maps::map('world', add = TRUE, col = 'red')
```

which doesn't look nice, but plotting the cells as polygons looks better:
```{r eval=EVAL}
plot(nit.c, breaks = "equal", reset = FALSE, axes = TRUE, as_points = FALSE, border = NA)
maps::map('world', add = TRUE, col = 'red')
```
