---
title: "spatiotemporal tidy arrays for R; first steps"
author: "Edzer Pebesma"
output:
  html_document:
    toc: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Spatiotemporal tidy arrays for R; first steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(collapse = TRUE)
```

```{r}
library(stars)
```

### Landsat image: GeoTIFF
```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
x = st_stars(tif)
str(x)
x
```

### Overviews
We can read rasters at a lower resolution when they contain so-called overviews. For this
GeoTIFF file, the were created with the `gdaladdo` utility, in particular
```
gdaladdo -r average L7_ETMs.tif  2 4 8 16
```
which adds course resolution versions by averaging blocks of pixels.

```{r}
x = st_stars(tif, options = c("OVERVIEW_LEVEL=1"))
image(x)
x = st_stars(tif, options = c("OVERVIEW_LEVEL=2"))
image(x)
x = st_stars(tif, options = c("OVERVIEW_LEVEL=3"))
image(x)
```

TODO: aspect ratio, title, values along x and y axis, degree symbols

### single NetCDF
```{r}
nc = system.file("nc/avhrr-only-v2.19810901.nc", package = "stars")
(x = st_stars(nc))
image(x)
library(units)
sapply(x, units)
sst = x[[1]]
hist(sst) # gives units, even if wrong
```

TODO: add time dimension (with n = 1); parse time from metadata

TODO: single NetCDF with multiple time and multiple depths

### multiple NetCDFs

join, based on time.

### dimensions to attributes and vice versa
