---
title: "spatiotemporal tidy arrays for R; first steps"
author: "Edzer Pebesma"
output:
  html_document:
    toc: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Spatiotemporal tidy arrays for R; first steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(collapse = TRUE)
```

```{r}
library(stars)
```

### Landsat image: GeoTIFF
```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
x = read_stars(tif)
str(x)
x
```

### Overviews
We can read rasters at a lower resolution when they contain so-called overviews. For this
GeoTIFF file, the were created with the `gdaladdo` utility, in particular
```
gdaladdo -r average L7_ETMs.tif  2 4 8 16
```
which adds course resolution versions by averaging blocks of pixels.

```{r}
(x = read_stars(tif, options = c("OVERVIEW_LEVEL=1")))
image(x)
(x = read_stars(tif, options = c("OVERVIEW_LEVEL=2")))
image(x)
(x = read_stars(tif, options = c("OVERVIEW_LEVEL=3")))
image(x)
```

TODO: aspect ratio, title, values along x and y axis, degree symbols

### `as.data.frame`

```{r}
names(x) = "value"
df = as.data.frame(x)
library(ggplot2)
library(viridis)
library(ggthemes)
ggplot() +  
  geom_tile(data=df, aes(x=x, y=y, fill=value), alpha=0.8) + 
  facet_wrap("band") +
  scale_fill_viridis() +
  coord_equal() +
  theme_map() +
  theme(legend.position="bottom") +
  theme(legend.key.width=unit(2, "cm"))
```

### `aperm`
change (permute) dimension order:
```{r}
aperm(x, c(1,3,2))
aperm(x, c("band", "x", "y"))
```

### single NetCDF

The sea surface temperatures (collected by PCMDI for
use by the IPCC) example file was downloaded from
https://www.unidata.ucar.edu/software/netcdf/examples/files.html

```{r}
nc = system.file("nc/tos_O1_2001-2002.nc", package = "stars")
(x = read_stars(nc))
plot(x, main = "")
```

### Raster with axes not aligned to N/E; conversion to `sf`
An example of a raster with rows and columns that are not aligned with E-W and N-S is
the following:

```{r}
geomatrix = system.file("tif/geomatrix.tif", package = "stars")
x = read_stars(geomatrix)
st_dimensions(x)
library(sf)
s = st_as_sf(x, as_points = FALSE)
par(cex.axis = .7) # font size axis tic labels 
plot(s, axes =TRUE, main = "geomatrix.tif", graticule = TRUE)
```


## conversion to/from raster:



TODO: single NetCDF with multiple time and multiple depths

TODO: conversion to and from raster

### multiple NetCDFs

DONE: join arrays, based on time, when list of nc files.
Done: image() breaks on non N/E aligned grid
Done: add time dimension (with n = 1); parse time from metadata

### dimensions to attributes and vice versa

`as.data.frame` converts dimensions to attributes, but all of them at once;
`st_as_sf` converts x/y into simple feature geometries, other things into attributes;
controlled, one-by one?

