---
title: '4. stars data model'
author: "Edzer Pebesma"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{4. stars data model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, dev = "png")
set.seed(13579)
```

This vignette explains the data model of `stars` objects, illustrated using artificial and real datasets.

## Stars objects

`stars` objects consist of 

* a (possibly empty) named list of arrays, each having named dimensions (`dim`) attribute
* an attribute called `dimensions` of class `dimensions` that carries dimension metadata
* a class name that includes `stars`

A `dimensions` object is a named list of `dimension` elements, each
describing the semantics a dimension of the data arrays (space,
time, type etc). In addition to that, a `dimensions` object has an
attribute called `raster` of class `stars_raster`, which is a named
list with three elements:

* `dimensions` length 2 character; the dimension names that constitute a spatial raster (or NA)
* `affine` length 2 numeric; the two affine parameters of the geotransform (or NA)
* `curvilinear` a boolean indicating whether a raster is a curvilinear raster (or NA)

The `affine` and `curvilinear` values are only relevant in case of
raster data, indicated by `dimensions` to have non-NA values.

A `dimension` object describes a _single_ dimension; it is a list with
named elements

* `from`: (numeric length 1): the start index of the array
* `to`: (numeric length 1): the end index of the array
* `offset`: (numeric length 1): the start coordinate (or time) value of the first pixel (i.e., a pixel/cell boundary)
* `delta`: (numeric length 1): the increment, or cell size
* `refsys`: (character, or `crs`): object describing the reference system; e.g. the PROJ string, or string `POSIXct` or `PCICt` (for 360 and 365 days/year calendars), or object of class `crs` (containing both EPSG code and proj4string)
* `point`: (logical length 1): boolean indicating whether cells/pixels refer to areas/periods, or to points/instances (may be NA)
* `values`: one of 
    * `NULL` (missing), 
    * a vector with coordinate values (numeric, `POSIXct`, `PCICt`, or `sfc`), 
    * an object of class `intervals` (a list with two vectors, `start` and `end`, with interval start- and end-values), or
    * a matrix with longitudes or latitudes for all cells (in case of curvilinear grids)

`from` and `to` will usually be 1 and the dimension size, but
`from` may be larger than 1 in case a sub-grid got was selected
(or cropped).

`offset` and `delta` only apply to _regularly_ discretized
dimensions, and are `NA` if this is not the case. If they are `NA`,
dimension values may be held in the `values` field.  Rectilinear and
curvilinear grids need grid values in `values` that can be either:

* for rectilinear grids: irregularly _spaced_ coordinate values, or coordinate _intervals_ of irregular width 
  (a rectilinear grid _can_ have one dimension that is regular), 
* for curvilinear grids: or a matrix with grid cell centre values for _all_ row/col combinations (usually in longitude or latitude).

Alternatively, `values` can contains a set of spatial geometries
encoded in an `sfc` vector ("list-column"), in which case we have a
[vector data cube](https://r-spatial.org/r/2022/09/12/vdc.html).

## Grid type

### Regular grids
With a very simple file created from a $4 \times 5$ matrix
```{r fig.width=4.5, fig.height=4}
suppressPackageStartupMessages(library(stars))
m = matrix(1:20, nrow = 5, ncol = 4)
dim(m) = c(x = 5, y = 4) # named dim
(s = st_as_stars(m))
```
we see that

* the rows (5) are mapped to the first dimension, the x-coordinate
* the columns (4) are mapped to the second dimension, the y-coordinate
* the `from` and `to` fields of each dimension define a range that corresponds to the array dimension:

```{r}
dim(s[[1]])
```
* offset and delta specify how increasing row and column index maps to x and y coordinate values respectively.

When we plot this object, using the `image` method for `stars` objects, 
```{r fig.width=4.5, fig.height=4}
image(s, text_values = TRUE, axes = TRUE)
```

we see that $(0,0)$ is the origin of the grid (grid corner), and $1$ the coordinate value increase from one index (row, col) to the next. It means that consecutive matrix columns represent grid lines, going from south to north. Grids defined this way are **regular**: grid cell size is constant everywhere.

Many actual grid datasets have y coordinates (grid rows) going from North to South (top to bottom); this is realised with a negative value for `delta`. We see that the grid origin $(0,0)$ did not change:
```{r fig.width=4.5, fig.height=4}
attr(s, "dimensions")[[2]]$delta = -1
image(s, text_values = TRUE, axes = TRUE)
```

An example is the GeoTIFF carried in the package, which, as probably all data sources read through GDAL, has a negative `delta` for the `y`-coordinate:
```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
st_dimensions(read_stars(tif))["y"]
```

### Raster attributes, rotated and sheared grids

Dimension tables of `stars` objects carry a `raster` attribute:
```{r}
str(attr(st_dimensions(s), "raster"))
```

which is a list that holds

* `dimensions`: character, the names of raster dimensions (if any), as opposed to e.g. spectral, temporal or other dimensions
* `affine`: numeric, the affine parameters
* `curvilinear`: a logical indicating whether the raster is curvilinear

These fields are needed at this level, because they describe properties of the array at a higher level than individual dimensions do: a pair of dimensions forms a raster, both `affine` and `curvilinear` describe how x and y _as a pair_ are derived from grid indexes (see below) when this cannot be done on a per-dimension basis.

With two affine parameters $a_1$ and $a_2$, $x$ and $y$ coordinates are derived from (1-based) grid indexes $i$ and $j$, grid offset values $o_x$ and $o_y$, and grid cell sizes $d_x$ and $d_y$ by

$$x = o_x + (i-1) d_x + (j-1) a_1$$
    
$$y = o_y + (i-1) a_2 + (j-1) d_y$$
Clearly, when $a_1=a_2=0$, $x$ and $y$ are entirely derived from their respective index, offset and cellsize.

Note that for integer indexes, the coordinates are that of the starting edge of a grid cell; to get the grid cell center of the top left grid cell (in case of a negative $d_y$), use $i=1.5$ and $j=1.5$.

We can rotate grids by setting $a_1$ and $a_2$ to a non-zero value:

```{r}
attr(attr(s, "dimensions"), "raster")$affine = c(0.1, 0.1)
plot(st_as_sf(s, as_points = FALSE), axes = TRUE, nbreaks = 20)
```

The rotation angle, in degrees, is

```{r}
atan2(0.1, 1) * 180 / pi
```

Sheared grids are obtained when the two rotation coefficients, $a_1$ and $a_2$, are unequal:
```{r}
attr(attr(s, "dimensions"), "raster")$affine = c(0.1, 0.2)
plot(st_as_sf(s, as_points = FALSE), axes = TRUE, nbreaks = 20)
```

Now, the y-axis and x-axis have different rotation in degrees of respectively
```{r}
atan2(c(0.1, 0.2), 1) * 180 / pi
```

## Rectilinear grids

[Rectilinear grids](https://en.wikipedia.org/wiki/Regular_grid) have orthogonal axes, but do not have congruent (equally sized and shaped) cells: each axis has its own irregular subdivision.

We can define a rectilinear grid by specifying the cell _boundaries_, meaning for every dimension we specify _one more_ value than the dimension size:
```{r}
x = c(0, 0.5, 1, 2, 4, 5)  # 6 numbers: boundaries!
y = c(0.3, 0.5, 1, 2, 2.2) # 5 numbers: boundaries!
(r = st_as_stars(list(m = m), dimensions = st_dimensions(x = x, y = y)))
st_bbox(r)
image(r, axes = TRUE, col = grey((1:20)/20))
```

Would we leave out the last value, than `stars` may come up with a _different_ cell boundary for the last cell, as this is now derived from the width of the one-but-last cell:
```{r}
x = c(0, 0.5, 1, 2, 4)  # 5 numbers: offsets only!
y = c(0.3, 0.5, 1, 2)   # 4 numbers: offsets only!
(r = st_as_stars(list(m = m), dimensions = st_dimensions(x = x, y = y)))
st_bbox(r)
```

This is not problematic if cells have a constant width, in which case the boundaries are reduced to an `offset` and `delta` value, irrespective whether an upper boundary is given:
```{r}
x = c(0, 1, 2, 3, 4)  # 5 numbers: offsets only!
y = c(0.5, 1, 1.5, 2)   # 4 numbers: offsets only!
(r = st_as_stars(list(m = m), dimensions = st_dimensions(x = x, y = y)))
st_bbox(r)
```

Alternatively, one can also set the _cell midpoints_ by specifying arguments `cell_midpoints` to the `st_dimensions` call:
```{r}
x = st_as_stars(matrix(1:9, 3, 3), 
                st_dimensions(x = c(1, 2, 3), y = c(2, 3, 10), cell_midpoints = TRUE))
```
When the dimension is regular, this results in `offset` being
shifted back with half a `delta`, or else in intervals derived from
the distances between cell centers.  This should obviously not be
done when cell boundaries are specified.

## Curvilinear grids

Curvilinear grids are grids whose grid lines are not straight. Rather than describing the curvature parametrically, the typical (HDF5 or NetCDF) files in which they are found have two raster layers with the longitudes and latitudes for every corresponding pixel of remaining layers. 

As an example, we will use a Sentinel 5P dataset available from package `starsdata`; this package can be installed with
```{r eval=FALSE}
install.packages("starsdata", repos = "http://pebesma.staff.ifgi.de", type = "source") 
```

The dataset is found here:
```{r}
(s5p = system.file("sentinel5p/S5P_NRTI_L2__NO2____20180717T120113_20180717T120613_03932_01_010002_20180717T125231.nc", package = "starsdata"))
```

```{r echo=FALSE}
EVAL = s5p != ""
```
We can construct the curvilinear `stars` raster by calling `read_stars` on the right sub-array:
```{r eval=EVAL}
subs = gdal_subdatasets(s5p)
subs[[6]]
```

For this array, we can see the GDAL metadata under item `GEOLOCATION`:
```{r eval=EVAL}
gdal_metadata(subs[[6]], "GEOLOCATION")
```

which reveals where, in this dataset, the longitude and latitude arrays are kept.

```{r eval=EVAL}
nit.c = read_stars(subs[[6]]) 
threshold = units::set_units(9e+36, mol/m^2)
nit.c[[1]][nit.c[[1]] > threshold] = NA
nit.c
```

The curvilinear array has the actual arrays (raster layers, matrices)
with longitude and latitude values read in its dimension table. We
can plot this file:

```{r eval=EVAL}
plot(nit.c, breaks = "equal", reset = FALSE, axes = TRUE, as_points = TRUE, 
		 pch = 16,  logz = TRUE, key.length = 1)
maps::map('world', add = TRUE, col = 'red')
```

```{r eval=EVAL}
plot(nit.c, breaks = "equal", reset = FALSE, axes = TRUE, as_points = FALSE, 
		 border = NA, logz = TRUE, key.length = 1)
maps::map('world', add = TRUE, col = 'red')
```

We can downsample the data by
```{r eval=EVAL}
(nit.c_ds = stars:::st_downsample(nit.c, 8))
plot(nit.c_ds, breaks = "equal", reset = FALSE, axes = TRUE, as_points = TRUE, 
		 pch = 16, logz = TRUE, key.length = 1)
maps::map('world', add = TRUE, col = 'red')
```

which doesn't look nice, but plotting the cells as polygons looks better:
```{r eval=EVAL}
plot(nit.c_ds, breaks = "equal", reset = FALSE, axes = TRUE, as_points = FALSE, 
		 border = NA, logz = TRUE, key.length = 1)
maps::map('world', add = TRUE, col = 'red')
```

Another approach would be to warp the curvilinear grid to a regular grid,
e.g. by

```{r eval=EVAL}
w = st_warp(nit.c, crs = 4326, cellsize = 0.25)
plot(w)
```
