---
title: "6. How `raster` functions map to `stars` functions"
author: "Sébastien Rochette, Adriano Fantini, Edzer Pebesma"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{6-raster-stars-mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, collapse = TRUE, comment = "#>")

suppressPackageStartupMessages(library(dplyr))
```

Here's an attempt at the table describing how `raster` functions map to `stars` functions, discussed in [issue #122](https://github.com/r-spatial/stars/issues/122). This table uses the functionality of the `raster` package as a template; it may be incomplete, imprecise or plain wrong, so take it with a pinch of salt. Any comment or correction is hugely appreciated, please contribute!

Some of the functions (`filter`, `slice`, `mutate`, `select`, `pull`, ...) are provided via `dplyr`, which must be loaded. See `?stars::dplyr`.

## COMMENT LEGEND

? = Not sure / unknown

\* = Not present, low priority

\# = Not present, high priority

NA = Not available by design

## Creating objects
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "{raster, stack, brick} (read)", "read_stars or read_stars(along = ...)","",
  "{stack, brick} (concatenate layers)", "c or c(along = ...)","",
  "subset", "{[ ] , slice, filter}","",
  "addLayer", "c() or c(along = ...)","",
  "dropLayer", "{[ ] , slice, filter}","",
  "unstack", "combine lapply and {[ ] , slice, filter}",""
) %>% 
  knitr::kable()
```


## Changing spatial extent and/or resolution of objects
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "merge", "c", "#, currently only works for adjacent objects",
  "mosaic", "st_mosaic", "",
  "crop", "filter, st_crop", "",
  "setExtent", "", "# maybe use st_warp?",
  "trim", "", "#",
  "aggregate", "aggregate", "WIP; raster's aggregate with fact=2 will not work, use st_warp in that case?",
  "disaggregate", "", "# use st_warp(use_gdal = TRUE)?",
  "resample", "{st_transform, st_warp}", "",
  "projectRaster", "{st_transform, st_warp}", "",
  "shift", "", "#, now use st_set_dimensions",
  "flip", "[] with reversed index", "#",
  "rotate", "", "*",
  "t", "", "NA"
) %>% 
  knitr::kable()
```

## Cell based computation
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "calc", "st_apply", "",
  "overlay", "c(along = , ...) %>% st_apply(...)", "",
  "cover", "[ ] <-", "",
  "mask", "[ ] <-", "",
  "cut", "cut", "",
  "subs", "", "",
  "reclassify", "", "* forcats::fct_recode ?",
  "init", "[ ] <-", "",
  "stackApply", "{[ ] , slice, filter} %>% st_apply", "",
  "stackSelect", "", "",
) %>% 
  knitr::kable()
```



## Spatial contextual computation
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "distance", "", "#",
  "gridDistance", "", "*",
  "distanceFromPoints", "", "#",
  "direction", "", "*",
  "focal", "f = st_apply(x1, 3, foc, w = matrix(1, 3, 3))", "See. [issue 176](https://github.com/r-spatial/stars/issues/176)",
  "localFun", "", "*",
  "boundaries", "st_as_sf(as_points=FALSE, merge=TRUE, connect8=TRUE)", "",
  "clump", "", "",
  "adjacent", "", "*",
  "area", "st_area", "",
  "terrain", "", "#",
  "Moran", "", ""
) %>% 
  knitr::kable()
```


## Model predictions
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "predict", "predict", "",
  "interpolate", "gstat::idw, gstat::krige", "`st_warp` has raster-raster interpolations of `gdalwarp`"
) %>% 
  knitr::kable()
```

## Data type conversion
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "rasterize", "st_as_stars", "",
  "rasterToPoints", "st_as_sf(as_points=TRUE)", "",
  "rasterToPolygons", "st_as_sf(as_points=FALSE, ...), st_polygonize", "",
  "rasterToContour", "st_contour", "requires GDAL >= 2.4.0",
  "rasterFromXYZ", "", "",
  "rasterFromCells", "", "",
) %>% 
  knitr::kable()
```

## Summarizing
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "cellStats", "st_apply", "",
  "summary", "print, summary(as.vector(. %>% pull))", "",
  "freq", "table", "*",
  "crosstab", "", "",
  "unique", "unique(as.vector(. %>% pull))", "",
  "zonal", "", "*"
) %>% 
  knitr::kable()
```


## Accessing values of objects
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "getValues", "{pull, [[ ]]}", "",
  "getValuesBlock", "{[ ] , slice, filter} %>% pull", "",
  "getValuesFocal", "{[ ] , slice, filter} %>% pull", "",
  "as.matrix", "[[ ]]", "currently behaves somewhat unexpectedly *", 
  "as.array", "[[ ]]", "currently behaves somewhat unexpectedly *", 
  "extract (by cell)", "{[ ] , slice, filter}", "",
  "extract (by polygon)", "x[sf_object]", "",
  "extract (by point)", "aggregate(stars_object, sf_object, function(x) x[1], as_points = FALSE)", "",
  "sampleRandom", "", "*", 
  "sampleRegular", "", "*",
  "minValue", "purrr::map(x, min)", "",
  "maxValue", "purrr::map(x, max)", "",
  "setMinMax", "", "," 
) %>% 
  knitr::kable()
```


## Plotting
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "plot", "plot, geom_stars", "",
  "plotRGB", "plot(x, rgb =...)", "",
  "spplot", "-", "",
  "image", "image", "",
  "persp", "-", "",
  "contour", "(st_contour, then sf::plot)", "",
  "filledContour", "(same)", "",
  "text", "text", "",
  "`hist`", "`hist(x[[1]])`", "",
  "barplot", "", "",
  "density", "", "",
  "pairs", "", "",
  "boxplot", "", ""
) %>% 
  knitr::kable()
```


## Getting and setting dimensions
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "ncol", "dim(x)[1]", "or use name instead of 1; cols may be the second dimension!", 
  "nrow", "dim(x)[2]", "or use name instead of 2; rows may be the first dimension!", 
  "ncell", "prod(dim(x))", "",
  "res", "st_dimensions", "can also not be a constant in case of rectilinear or curvilinear grids", 
  "nlayers", "-", "there is no concept of layers in stars", 
  "names", "names", "",
  "xres", "st_dimensions, look for delta", "may not be a constant in case of rectilinear or curvilinear grids", 
  "yres", "st_dimensions, look for delta", "may not be a constant in case of rectilinear or curvilinear grids", 
  "xmin", "st_bbox(x)[1]", "",
  "xmax", "st_bbox(x)[3]", "",
  "ymin", "st_bbox(x)[2]", "",
  "ymax", "st_bbox(x)[4]", "",
  "extent", "st_bbox(x)", "different ordering of numbers", 
  "origin", "-", "",
  "projection", "st_crs(x)", "",
  "isLonLat", "st_is_longlat(st_crs(x))", "",
  "filename", "", "stars_proxy objects carry file names where otherwise the array data is", 
  "bandnr", "", "stars has no general concept of bands", 
  "nbands", "dim(x)[3]", "may also be time; bands may also be in another dimension, or have another name", 
  "compareRaster", "all.equal(st_dimensions(x), st_dimensions(y))", "*", 
  "NAvalue", "", "-"
) %>% 
  knitr::kable()
```


## Computing row, column, cell numbers and coordinates
```{r}
tribble(
  ~raster, ~stars, ~`Note/comment`,
  "xFromCol", "st_get_dimension_values(., 'x')[col]", "I am not sure how to generally get the x dimension name - here it is x",
  "yFromRow", "st_get_dimension_values(., 'y')[row]", "I am not sure how to generally get the y dimension name - here it is y", 
  "xFromCell", "", "",
  "yFromCell", "", "",
  "xyFromCell", "", "", 
  "colFromX", "", "*",
  "rowFromY", "", "*", 
  "rowColFromCell", "", "",
  "cellFromXY", "", "",
  "cellFromRowCol", "", "",
  "cellsFromExtent", "", "",
  "coordinates", "st_coordinates", "",
  "validCell", "", "",
  "validCol", "col %>% between(st_dimensions(.)$x$from, st_dimensions(.)$x$to)", "raster columns are not always named 'x'", 
  "validRow", "row %>% between(st_dimensions(.)$y$from, st_dimensions(.)$y$to)", "raster columns are not always named 'y'", 
  "setValues", "[ ] <-", "",
  "writeRaster", "write_stars", "currently uses GDAL, somewhat limited", 
  "KML", "", ""
) %>% 
  knitr::kable()
```


The format of this table follows the `raster-package` entry in the [raster manual](https://cran.r-project.org/web/packages/raster/raster.pdf).
