---
title: "5. vector-raster conversions, reprojection, warping"
author: "Edzer Pebesma"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{5. vector-raster and raster-vector conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(collapse = TRUE)
suppressPackageStartupMessages(library(sf))
ev = sf_extSoftVersion()["GDAL"] >= "2.4.0"
```

This vignette shows how `stars` object can be moved from vector
and raster representations.

# Rasterizing an `sf` vector object

```{r}
library(stars)
system.file("gpkg/nc.gpkg", package = "sf") %>%
  read_sf() %>%
  st_transform(32119) -> nc
nc$dens = nc$BIR79 / units::set_units(st_area(nc), km^2)
(nc.st = st_rasterize(nc["dens"], deltax = 5000, deltay = 5000))
plot(nc.st)
```

The algorithm used is the GDAL `rasterize` utility, all options of
this utility can be passed to `st_rasterize`. The geometry of the
final raster can be controlled by passing a target bounding box and
either the raster dimensions `nx` and `ny`, or pixel size by the
`deltax` and `deltay` parameters.

# Vectorizing a raster object to an `sf` object

`stars` objects can be converted into an `sf` object using
`st_as_sf`.  It has a number of options, depending on whether
pixels represent the point value at the pixel center, or small
square polygons with a single value.

We will work again with the landsat-7 6-band image, but will select
the first band and round the values:
```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
x = read_stars(tif)[, 1:50, 1:50, 1:2]
x[[1]] = round(x[[1]]/5)
```

## Polygonizing

In case raster cells reflect point values and we want to get
a vector representatoin of the whole field, we can draw contour
lines and export the contour sets (only available when the 
GDAL version is at least 2.4.0):
```{r eval=ev}
#p = st_as_sf(x, as_points = TRUE, merge = TRUE)
#plot(p)
```

## Exporting to points

Alternatively, we can simply export all the pixels as points, and
get them either as a wide table with all bands per point, and no
replicated `POINT` geometries:
```{r}
st_as_sf(x, as_points = TRUE, merge = FALSE)
```
or as a long table with a single attribute and all points replicated:
```{r}
st_as_sf(x, as_points = TRUE, merge = FALSE, long = TRUE)
```
as we can see, an additional attribute `band` now indicates which band is concerned.

## Exporting to polygons

Alternatively, we can export to polygons and either get a single polygon per
pixel, as in
```{r}
st_as_sf(x[1], as_points = FALSE, merge = FALSE)
```
or merge polygons that have identical pixel values;
```{r}
p = st_as_sf(x, as_points = FALSE, merge = TRUE)
```
When plotted with boundaries, we see the resolved boundaries of areas with the
same pixel value:
```{r}
plot(p)
```

A further option `connect8` can be set to `TRUE` to use 8
connectedness, rather than the default 4 connectedness algorithm.
In both cases, the polygons returned will often be invalid according
to the simple feature standard, but can be made valid using
`lwgeom::st_make_valid`.

# Switching between vector and raster in `stars` objects

We can convert a raster dimension to a vector dimension while keeping other dimensions as they are in a `stars` object by
```{r}
x.sf = st_xy2sfc(x, as_points = TRUE)
x.sf
```
which also requires setting the `as_points` arguments as in `st_as_sf`.

# Reprojecting a raster

If we accept that curvilinear rasters are rasters too, and that
regular and rectilinear grids are special cases of curvilinear grids,
reprojecting a raster is no longer a "problem", it just recomputes
new coordinates for every raster cell, and generally results in a
curvilinear grid (that sometimes can be brought back to a regular
or rectilinear grid). If curvilinear grid cells are represented
by coordinates of the cell center, the actual shape of a grid cell
gets lost, and this may be a larger effect if grid cells are large
or if the transformation is stronger non-linear.

An example of the reprojection of the grid created above is
```{r}
nc.st %>% st_transform("+proj=laea +lat_0=34 +lon_0=-60") -> nc.curv
nc.curv
plot(nc.curv, border = NA, graticule = TRUE)
```

where it should be noted that the dimensionality of the grid didn't
change: the same set of raster cells has been replotted in the new
CRS, but now in a curvilinear grid.

# Warping a raster

Warping a raster means creating a new _regular_ grid in a new CRS,
based on a (usually regular) grid in another CRS. We can do the
transformation of the previous section by first creating a target
grid:
```{r}
nc %>% st_transform("+proj=laea +lat_0=34 +lon_0=-60") %>% st_bbox() %>%
	st_as_stars() -> newgrid
```
and then warping the old raster to the new
```{r}
nc.st %>% st_warp(newgrid) -> nc.new
nc.new 
plot(nc.new)
```

This new object has a regular grid in the new CRS, aligned with
the new x- and y-axes.
