---
title: "8. NetCDF Proxy Workflows"
author: "David Blodgett"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{8. NetCDF Proxy Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(dplyr))
EVAL = suppressWarnings(require(starsdata, quietly = TRUE))
```

NetCDF data sources are available via more _and_ less granular files and/or OPeNDAP endpoints. This article demonstrates how `stars` enables discovery, access, and processing of NetCDF data across a wide range of such source-data organization schemes.

We'll start with some basics using datasets included with the `stars` installation. A call to `read_ncdf()`, for a dataset smaller than the default threshold, will just read in all the data. Below we read in and display the `reduced.nc` NetCDF file.

```{r}
library(stars)
f <- system.file("nc/reduced.nc", package = "stars")
(nc <- read_ncdf(f))
```

Let's assume `reduced.nc` was 10 years of hourly data, rather than 1 time step. It would be over 10GB rather than about 130KB and we would not be able to just read it all into memory. In this case, we need a way to read the file's metadata such that we could iterate over it in a way that meets the needs of our workflow objectives. This is where `proxy = TRUE` comes in. Below, we'll lower the option that controls whether `read_ncdf()` defaults to proxy _and_ use `proxy = TRUE` to show both ways of getting the same result.

```{r}
old_options <- options("stars.n_proxy" = 100)
(nc <- read_ncdf(f, proxy = TRUE))
options(old_options)
```

The above shows that we have a NetCDF sourced stars proxy derived from the `reduced.nc` file. We see it has four variables and their units are displayed. The normal `stars` `dimension(s)` are available and a `nc_request` object is also available. The `nc_request` object contains the information needed to make requests for data according to the dimensions of the NetCDF data source. With this information, we have what we need to request a chunk of data that is what we want and not too large.

```{r}

(nc <- read_ncdf(f, 
				 var = "sst", 
				 ncsub = cbind(start = c(90, 45, 1 , 1), 
				 			  count = c(90, 45, 1, 1))))

plot(nc)
```
